---
title: "Part_2"
format: html
editor: visual
---

```{r}
#| echo: FALSE
#| warnings: FALSE
#| message: FALSE
# install packages if necessary
#install.packages("caret")
#install.packages("lattice")
#install.packages("recipes")

```

# Load data

```{r}
#| echo: TRUE
#| echo: TRUE
#| warnings: FALSE
#| message: FALSE
library(tidyverse)
library(rsample)
library(caret)

## 1. Read fresh
smart1 <- read.csv2("smartc.csv")

## 2. Convert the true categorical variables to factors FIRST
smart1 <- smart1 %>%
  mutate(
    EVENT    = factor(EVENT,    levels = c(0,1),   labels = c("no","yes")),
    DIABETES = factor(DIABETES, levels = c(0,1),   labels = c("no","yes")),
    CEREBRAL = factor(CEREBRAL, levels = c(0,1),   labels = c("no","yes")),
    CARDIAC  = factor(CARDIAC,  levels = c(0,1),   labels = c("no","yes")),
    AAA      = factor(AAA,      levels = c(0,1),   labels = c("no","yes")),
    PERIPH   = factor(PERIPH,   levels = c(0,1),   labels = c("no","yes")),
    STENOSIS = factor(STENOSIS, levels = c(0,1),   labels = c("no","yes")),
    SEX      = factor(SEX,      levels = c(1,2),   labels = c("male","female")),
    SMOKING  = factor(SMOKING,  levels = c(1,2,3), labels = c("never","former","current")),
    alcohol  = factor(alcohol,  levels = c(1,2,3), labels = c("never","former","current")),
    albumin  = factor(albumin,  levels = c(1,2,3), labels = c("no","micro","macro"))
  ) %>%
  ## 3. Decimal columns read as text -> numeric. The factors above are left untouched.
  mutate(across(where(is.character), as.numeric))

## 4. Sanity checks
stopifnot(is.numeric(smart1$BMI), sum(is.na(smart1$EVENT)) == 0)
str(smart1[c("EVENT", "BMI", "HDL", "IMT", "albumin")])  # EVENT/albumin factor; BMI/HDL/IMT num

## 5. Split + fit (same model as 1c)
set.seed(2026)
smart_split <- initial_split(smart1, prop = 0.75, strata = EVENT)
train_data  <- training(smart_split)
test_data   <- testing(smart_split)

fit <- glm(EVENT ~ AGE + SEX + BMI + SYSTH + HDL + DIABETES + HISTCAR2 +
             HOMOC + log(CREAT) + STENOSIS + IMT + SMOKING + alcohol + albumin,
           data = train_data, family = binomial)
```

# Logistic Regression Model

```{r}
#| echo: TRUE
#| warnings: FALSE
#| message: FALSE

model <- glm(EVENT ~ AGE + SEX + BMI + SYSTH + HDL + DIABETES + HISTCAR2 +
               HOMOC + log(CREAT) + STENOSIS + IMT + SMOKING + alcohol + albumin,
             data = smart1, family = binomial)

```

# Question A

The performance of the logistic regression model is evaluated using 5-folds cross validation (n = 3873). At the default probability threshold of 0.5, the mean sensitivity was 0.05 (SD = 0.02) and the mean specificity was 0.99 (SD = 0.05). The mean positive predictive value is 0.45, meaning that 45% of positive predictions were correct. The mean negative predictive value is 0.89, meaning that 89% of negative predictions were correct. However, the PPV and NPV are strongly influenced by the prevalence in this cohort (approximately 12%), a higher prevalence would expected to yield a higher PPV and a lower NPV even with the same underlying model performance.

The mean area under the ROC curve across the folds was 0.73 (SD = 0.03), and the AUC of the ROC based on the pooled cross-validated predictions was 0.73, see Figure. This means that a randomly chosen patient with an event receives a higher predicted probability than a randomly chosen patient without an event in about 73% of cases. This is better than the random classifier (AUC = 0.5), but far from perfect discrimination (AUC = 1). Furthermore, the ROC curve shows how sensitivity and specificity trade off as the threshold varies. For example, for a sensitivity of roughly 0.8, a specificity of roughly 0.5 should be accepted.

```{r}
#| echo: TRUE
#| warnings: FALSE
#| message: FALSE

# same proportion in every fold, check if that is correct
# way of validating the final model of part 1 with no external data available

library(caret)
library(pROC)

set.seed(1234)

folds <- createFolds(smart1$EVENT, k = 5, returnTrain = TRUE) # ensures by default that EVENT proportions are equal 
print(str(folds))

train_control <- trainControl(
  method = "cv",
  number = 5,
  index = folds,
  classProbs = TRUE,                  # needed to get predicted probabilities, not just class labels
  summaryFunction = twoClassSummary,  # computes ROC/AUC, sensitivity, specificity
  savePredictions = "final"           # keeps held-out predictions for plotting the ROC curve
)

model_glm <- train(
  EVENT ~ AGE + SEX + BMI + SYSTH + HDL + DIABETES + HISTCAR2 +
    HOMOC + log(CREAT) + STENOSIS + IMT + SMOKING + alcohol + albumin,
  data = smart1,
  method = "glm",
  family = "binomial",
  trControl = train_control,
  metric = "ROC"                      # tells caret to select/report based on AUC
)

# View CV performance
print(model_glm)
print(model_glm$results)     # mean ROC (AUC), Sens, Spec across the 5 fold (sensitivity and specificity are flipped due to sequence in labeling)

# Extract the saved CV predictions
cv_preds <- model_glm$pred

# Confusion matrix at 0.5, using the labels caret stored
confusionMatrix(cv_preds$pred, cv_preds$obs, positive = "yes")  # sensitivity for events

# Build ROC object (probability of "yes" vs actual class)
roc_obj <- roc(response = cv_preds$obs, 
               predictor = cv_preds$yes,
               levels = c("no", "yes"))

# Plot
plot(roc_obj, 
     main = "ROC Curve (5-fold CV)", 
     col = "blue", 
     lwd = 2,
     xlim = c(1, 0),
     xaxs = "i",          
     yaxs = "i",
     xlab = "Specificity",
     ylab = "Sensitivity")

auc(roc_obj)
```

# Question B

```{r}
#| echo: TRUE

# Calibration 

# Predicted P(EVENT = "yes") on the test set + observed class
testProbs <- tibble(
  .pred_yes = predict(fit, newdata = test_data, type = "response"),
  Class     = test_data$EVENT
)

# Calibration plot on the test data (caret, slide 31)
cal_obj <- caret::calibration(Class ~ .pred_yes, data = testProbs,
                              class = "yes", cuts = 10)
ggplot(cal_obj) +
  labs(title = "Calibration plot – test data",
       x = "Bin midpoint (predicted probability)",
       y = "Observed event percentage")
```

```{r}
# Cross-validated calibration with intercept & slope
library(rms)

form <- EVENT ~ AGE + SEX + BMI + SYSTH + HDL + DIABETES + HISTCAR2 +
  HOMOC + log(CREAT) + STENOSIS + IMT + SMOKING + alcohol + albumin

set.seed(2026)
K      <- 10
folds  <- sample(rep(1:K, length.out = nrow(smart1)))
cvprob <- rep(NA_real_, nrow(smart1))

for (k in 1:K) {
  m <- glm(form, data = smart1[folds != k, ], family = binomial)
  cvprob[folds == k] <- predict(m, newdata = smart1[folds == k, ], type = "response")
}

# val.prob() needs the outcome as 0/1: factor "no"/"yes" -> 1/2 -> 0/1
y01 <- as.numeric(smart1$EVENT) - 1

cal_stats <- val.prob(cvprob, y01)   # draws the plot AND returns the statistics
title("Cross-validated calibration plot (10-fold)")

# Pull out the numbers you report in the text:
round(cal_stats[c("Intercept", "Slope", "C (ROC)", "Brier")], 3)
```

# Question C

```{r}
#| echo: TRUE
```
