---
title: "Part_2"
format: html
editor: visual
---

```{r}
#| echo: FALSE
#| warnings: FALSE
#| message: FALSE
# install packages if necessary
#install.packages("caret")
#install.packages("lattice")
#install.packages("recipes")

```

# Load data

```{r}
#| echo: TRUE
#| warnings: FALSE
#| message: FALSE

library(tidyverse)

#load data
smart1 <- read.csv2("smartc.csv")

smart1 <- smart1 %>%
  mutate(
    # Binary yes/no variables
    EVENT     = factor(EVENT,     levels = c(0, 1), labels = c("no", "yes")),
    DIABETES  = factor(DIABETES,  levels = c(0, 1), labels = c("no", "yes")),
    CEREBRAL  = factor(CEREBRAL,  levels = c(0, 1), labels = c("no", "yes")),
    CARDIAC   = factor(CARDIAC,   levels = c(0, 1), labels = c("no", "yes")),
    AAA       = factor(AAA,       levels = c(0, 1), labels = c("no", "yes")),
    PERIPH    = factor(PERIPH,    levels = c(0, 1), labels = c("no", "yes")),
    STENOSIS  = factor(STENOSIS,  levels = c(0, 1), labels = c("no", "yes")),
    
    # Binary but coded 1/2
    SEX       = factor(SEX,       levels = c(1, 2), labels = c("male", "female")),
    
    # Multi-level nominal/ordinal variables
    SMOKING   = factor(SMOKING,   levels = c(1, 2, 3), labels = c("never", "former", "current")),
    alcohol   = factor(alcohol,   levels = c(1, 2, 3), labels = c("never", "former", "current")),
    albumin   = factor(albumin, levels = c(1, 2, 3),
                       labels = c("no", "micro", "macro"),
                       ordered = TRUE))

#change data types to numeric if necessary
smart1 <- smart1 %>%
  mutate(across(where(is.character), as.numeric))

```

# Logistic Regression Model

```{r}
#| echo: TRUE
#| warnings: FALSE
#| message: FALSE

model <- glm(EVENT ~ AGE + SEX + BMI + SYSTH + HDL + DIABETES + HISTCAR2 +
               HOMOC + log(CREAT) + STENOSIS + IMT + SMOKING + alcohol + albumin,
             data = smart1, family = binomial)

```

# Question A

The performance of the logistic regression model is evaluated using 5-folds cross validation (n = 3873). At the default probability threshold of 0.5, the mean sensitivity was 0.05 (SD = 0.02) and the mean specificity was 0.99 (SD = 0.05). The mean positive predictive value is 0.45, meaning that 45% of positive predictions were correct. The mean negative predictive value is 0.89, meaning that 89% of negative predictions were correct. However, the PPV and NPV are strongly influenced by the prevalence in this cohort (approximately 12%), a higher prevalence would expected to yield a higher PPV and a lower NPV even with the same underlying model performance.

The mean area under the ROC curve across the folds was 0.73 (SD = 0.03), and the AUC of the ROC based on the pooled cross-validated predictions was 0.73, see Figure. This means that a randomly chosen patient with an event receives a higher predicted probability than a randomly chosen patient without an event in about 73% of cases. This is better than the random classifier (AUC = 0.5), but far from perfect discrimination (AUC = 1). Furthermore, the ROC curve shows how sensitivity and specificity trade off as the threshold varies. For example, for a sensitivity of roughly 0.8, a specificity of roughly 0.5 should be accepted.

```{r}
#| echo: TRUE
#| warnings: FALSE
#| message: FALSE

# same proportion in every fold, check if that is correct
# way of validating the final model of part 1 with no external data available

library(caret)
library(pROC)

set.seed(1234)

folds <- createFolds(smart1$EVENT, k = 5, returnTrain = TRUE) # ensures by default that EVENT proportions are equal 
print(str(folds))

train_control <- trainControl(
  method = "cv",
  number = 5,
  index = folds,
  classProbs = TRUE,                  # needed to get predicted probabilities, not just class labels
  summaryFunction = twoClassSummary,  # computes ROC/AUC, sensitivity, specificity
  savePredictions = "final"           # keeps held-out predictions for plotting the ROC curve
)

model_glm <- train(
  EVENT ~ AGE + SEX + BMI + SYSTH + HDL + DIABETES + HISTCAR2 +
    HOMOC + log(CREAT) + STENOSIS + IMT + SMOKING + alcohol + albumin,
  data = smart1,
  method = "glm",
  family = "binomial",
  trControl = train_control,
  metric = "ROC"                      # tells caret to select/report based on AUC
)

# View CV performance
print(model_glm)
print(model_glm$results)     # mean ROC (AUC), Sens, Spec across the 5 fold (sensitivity and specificity are flipped due to sequence in labeling)

# Extract the saved CV predictions
cv_preds <- model_glm$pred

# Confusion matrix at 0.5, using the labels caret stored
confusionMatrix(cv_preds$pred, cv_preds$obs, positive = "yes")  # sensitivity for events

# Build ROC object (probability of "yes" vs actual class)
roc_obj <- roc(response = cv_preds$obs, 
               predictor = cv_preds$yes,
               levels = c("no", "yes"))

# Plot
plot(roc_obj, 
     main = "ROC Curve (5-fold CV)", 
     col = "blue", 
     lwd = 2,
     xlim = c(1, 0),
     xaxs = "i",          
     yaxs = "i",
     xlab = "Specificity",
     ylab = "Sensitivity")

auc(roc_obj)
```

# Question B

```{r}
#| echo: TRUE
```

# Question C

```{r}
#| echo: TRUE
```
