---
title: "Part_2"
format: html
editor: visual
---

```{r}
#| echo: FALSE
#| warnings: FALSE
#| message: FALSE
# install packages if necessary
#install.packages("caret")
#install.packages("lattice")
#install.packages("recipes")

```

# Load data

```{r}
#| echo: TRUE
#| echo: TRUE
#| warnings: FALSE
#| message: FALSE
library(tidyverse)
library(rsample)
library(caret)
library(rms)

## 1. Read fresh
smart1 <- read.csv2("smartc.csv")

## 2. Convert the true categorical variables to factors FIRST
smart1 <- smart1 %>%
  mutate(
    EVENT    = factor(EVENT,    levels = c(0,1),   labels = c("no","yes")),
    DIABETES = factor(DIABETES, levels = c(0,1),   labels = c("no","yes")),
    CEREBRAL = factor(CEREBRAL, levels = c(0,1),   labels = c("no","yes")),
    CARDIAC  = factor(CARDIAC,  levels = c(0,1),   labels = c("no","yes")),
    AAA      = factor(AAA,      levels = c(0,1),   labels = c("no","yes")),
    PERIPH   = factor(PERIPH,   levels = c(0,1),   labels = c("no","yes")),
    STENOSIS = factor(STENOSIS, levels = c(0,1),   labels = c("no","yes")),
    SEX      = factor(SEX,      levels = c(1,2),   labels = c("male","female")),
    SMOKING  = factor(SMOKING,  levels = c(1,2,3), labels = c("never","former","current")),
    alcohol  = factor(alcohol,  levels = c(1,2,3), labels = c("never","former","current")),
    albumin  = factor(albumin,  levels = c(1,2,3), labels = c("no","micro","macro"))
  ) %>%
  ## 3. Decimal columns read as text -> numeric. The factors above are left untouched.
  mutate(across(where(is.character), as.numeric))

## 4. Sanity checks
stopifnot(is.numeric(smart1$BMI), sum(is.na(smart1$EVENT)) == 0)
str(smart1[c("EVENT", "BMI", "HDL", "IMT", "albumin")])  # EVENT/albumin factor; BMI/HDL/IMT num

## 5. Split + fit (same model as 1c)
set.seed(2026)
smart_split <- initial_split(smart1, prop = 0.75, strata = EVENT)
train_data  <- training(smart_split)
test_data   <- testing(smart_split)

fit <- glm(EVENT ~ AGE + SEX + BMI + SYSTH + HDL + DIABETES + HISTCAR2 +
             HOMOC + log(CREAT) + STENOSIS + IMT + SMOKING + alcohol + albumin,
           data = train_data, family = binomial)
```

# Logistic Regression Model

```{r}
#| echo: TRUE
#| warnings: FALSE
#| message: FALSE

model <- glm(EVENT ~ AGE + SEX + BMI + SYSTH + HDL + DIABETES + HISTCAR2 +
               HOMOC + log(CREAT) + STENOSIS + IMT + SMOKING + alcohol + albumin,
             data = smart1, family = binomial)

```

# Question A

The performance of the logistic regression model is evaluated using 5-folds cross validation (n = 3873). At the default probability threshold of 0.5, the mean sensitivity was 0.05 (SD = 0.02) and the mean specificity was 0.99 (SD = 0.005). The mean positive predictive value is 0.45, meaning that 45% of positive predictions were correct. The mean negative predictive value is 0.89, meaning that 89% of negative predictions were correct. However, the PPV and NPV are strongly influenced by the prevalence in this cohort (approximately 12%), a higher prevalence would expected to yield a higher PPV and a lower NPV even with the same underlying model performance.

The mean area under the ROC curve across the folds was 0.73 (SD = 0.03), and the AUC of the ROC based on the pooled cross-validated predictions was 0.73, see Figure. This means that a randomly chosen patient with an event receives a higher predicted probability than a randomly chosen patient without an event in about 73% of cases. This is better than the random classifier (AUC = 0.5), but far from perfect discrimination (AUC = 1). Furthermore, the ROC curve shows how sensitivity and specificity trade off as the threshold varies. For example, for a sensitivity of roughly 0.8, a specificity of roughly 0.5 should be accepted.

```{r}
#| echo: TRUE
#| warnings: FALSE
#| message: FALSE

# same proportion in every fold, check if that is correct
# way of validating the final model of part 1 with no external data available

library(caret)
library(pROC)

set.seed(1234)

folds <- createFolds(smart1$EVENT, k = 5, returnTrain = TRUE) # ensures by default that EVENT proportions are equal 
print(str(folds))

train_control <- trainControl(
  method = "cv",
  number = 5,
  index = folds,
  classProbs = TRUE,                  # needed to get predicted probabilities, not just class labels
  summaryFunction = twoClassSummary,  # computes ROC/AUC, sensitivity, specificity
  savePredictions = "final"           # keeps held-out predictions for plotting the ROC curve
)

model_glm <- train(
  EVENT ~ AGE + SEX + BMI + SYSTH + HDL + DIABETES + HISTCAR2 +
    HOMOC + log(CREAT) + STENOSIS + IMT + SMOKING + alcohol + albumin,
  data = smart1,
  method = "glm",
  family = "binomial",
  trControl = train_control,
  metric = "ROC"                      # tells caret to select/report based on AUC
)

# View CV performance
print(model_glm)
print(model_glm$results)     # mean ROC (AUC), Sens, Spec across the 5 fold (sensitivity and specificity are flipped due to sequence in labeling)

# Extract the saved CV predictions
cv_preds <- model_glm$pred

# Confusion matrix at 0.5, using the labels caret stored
confusionMatrix(cv_preds$pred, cv_preds$obs, positive = "yes")  # sensitivity for events

# Build ROC object (probability of "yes" vs actual class)
roc_obj <- roc(response = cv_preds$obs, 
               predictor = cv_preds$yes,
               levels = c("no", "yes"))

# Plot
plot(roc_obj, 
     main = "ROC Curve (5-fold CV)", 
     col = "blue", 
     lwd = 2,
     xlim = c(1, 0),
     xaxs = "i",          
     yaxs = "i",
     xlab = "Specificity",
     ylab = "Sensitivity")

auc(roc_obj)
```

# Question B


```{r}
#| echo: TRUE

# Calibration 

# Predicted P(EVENT = "yes") on the test set + observed class
testProbs <- tibble(
  .pred_yes = predict(fit, newdata = test_data, type = "response"),
  Class     = test_data$EVENT
)

# Calibration plot on the test data (caret, slide 31)
cal_obj <- caret::calibration(Class ~ .pred_yes, data = testProbs,
                              class = "yes", cuts = 10)
ggplot(cal_obj) +
  labs(title = "Calibration plot – test data",
       x = "Bin midpoint (predicted probability)",
       y = "Observed event percentage")
   cal_obj$data[, c("bin", "Percent", "Count", "midpoint")]
```
The calibration of the logistic regression model is evaluated using two calibration plots. First, the model is fitted on the training data (75%) and the predicted probabilities are calculated for the test data (25%). The test patients are divided into 10 bins based on their predicted risk, and for each bin the observed event percentage is plotted with a 95% confidence interval (see Figure). If the model is well calibrated, the points lie on the diagonal. In the first two bins (0–10% and 10–20%), which contain most of the patients, the observed event percentages are about 7% and 11%. These are close to the diagonal with narrow confidence intervals, meaning that for low-risk patients the predicted risks agree with the observed event rates. The bins at 25% and 35% are also close to the diagonal. At 45% and 55% the observed percentages (28% and 40%) are lower than predicted, but the confidence intervals are wide and still include the diagonal. Above 60% the bins contain only one or two patients (confidence intervals from 0 to 97.5%), and the 95% bin is empty, which our caret shows as 0%. These bins give no information about the calibration of the model. However, because the test set is small, the calibration for high-risk patients cannot be judged from this plot.Therefore, we also evaluate using 5-fold cross-validation on the full data set (n = 3873), in the same way as in question A

```{r}
# Cross-validated calibration with intercept & slope

form <- EVENT ~ AGE + SEX + BMI + SYSTH + HDL + DIABETES + HISTCAR2 +
  HOMOC + log(CREAT) + STENOSIS + IMT + SMOKING + alcohol + albumin

set.seed(2026)
K      <- 5
folds  <- sample(rep(1:K, length.out = nrow(smart1)))
cvprob <- rep(NA_real_, nrow(smart1))

for (k in 1:K) {
  m <- glm(form, data = smart1[folds != k, ], family = binomial)
  cvprob[folds == k] <- predict(m, newdata = smart1[folds == k, ], type = "response")
}

# val.prob() needs the outcome as 0/1: factor "no"/"yes" -> 1/2 -> 0/1
y01 <- as.numeric(smart1$EVENT) - 1

cal_stats <- val.prob(cvprob, y01)   # draws the plot AND returns the statistics
title("Cross-validated calibration plot (5-fold)")

# Pull out the numbers you report in the text:
round(cal_stats[c("Intercept", "Slope", "C (ROC)", "Brier")], 3)
```
The calibration slope is 0.925 and the calibration intercept is -0.131. For good calibration the slope should be close to 1 and the intercept close to 0. A slope slightly below 1 means that the predictions are somewhat too extreme, mainly for high predicted risks, which indicates a little overfitting. This can be expected with 14 predictors and no variable selection. The intercept is estimated together with the slope and is therefore hard to interpret on its own. The mean absolute difference between predicted and observed risk (Eavg) is 0.005, meaning that on average the predictions agree with the observed outcomes. The Spiegelhalter test also shows no significant miscalibration (z = 0.556, p = 0.578). Confidence intervals for the intercept and slope were not calculated, but these could be obtained with bootstrap resampling.

The histogram at the bottom of the plot shows that almost all patients have a predicted risk below 0.2, which is in line with the prevalence of around 12%. In this range the nonparametric curve follows the ideal line. Above a predicted risk of about 0.45 the curve falls below the ideal line. For example, for a predicted risk of about 0.85 the observed risk is only about 0.7. The maximum difference between predicted and observed risk (Emax) is 0.149 and lies in this high-risk range. This means that the model overestimates the risk for the small group of high-risk patients, which is in line with the slope below 1. The C-statistic of 0.736 is similar to the AUC from question A (0.73), and the Brier score is 0.095.

In conclusion, the model is well calibrated for the large majority of patients with a predicted risk below 0.2 to 0.3, but overestimates the risk for the few patients with a high predicted risk.

# Question C

```{r}
#| echo: TRUE
```
