---
title: "Part_2 Group 6"
format: html
editor: visual
---

Authors: Hilde Gelinck, Niels Pleijsier and Jesse de Heus

```{r}
#| echo: FALSE
#| warnings: FALSE
# install packages if necessary
#install.packages()
#install.packages("caret")
#install.packages("lattice")
#install.packages("recipes")

```

# Load data

```{r}
#| echo: TRUE
#| warnings: FALSE

library(tidyverse)

#load data
smart1 <- read.csv2("smartc.csv")

smart1 <- smart1 %>%
  mutate(
    # Binary yes/no variables
    EVENT     = factor(EVENT,     levels = c(0, 1), labels = c("no", "yes")),
    DIABETES  = factor(DIABETES,  levels = c(0, 1), labels = c("no", "yes")),
    CEREBRAL  = factor(CEREBRAL,  levels = c(0, 1), labels = c("no", "yes")),
    CARDIAC   = factor(CARDIAC,   levels = c(0, 1), labels = c("no", "yes")),
    AAA       = factor(AAA,       levels = c(0, 1), labels = c("no", "yes")),
    PERIPH    = factor(PERIPH,    levels = c(0, 1), labels = c("no", "yes")),
    STENOSIS  = factor(STENOSIS,  levels = c(0, 1), labels = c("no", "yes")),
    
    # Binary but coded 1/2
    SEX       = factor(SEX,       levels = c(1, 2), labels = c("male", "female")),
    
    # Multi-level nominal/ordinal variables
    SMOKING   = factor(SMOKING,   levels = c(1, 2, 3), labels = c("never", "former", "current")),
    alcohol   = factor(alcohol,   levels = c(1, 2, 3), labels = c("never", "former", "current")),
    albumin   = factor(albumin, levels = c(1, 2, 3),
                       labels = c("no", "micro", "macro"),
                       ordered = TRUE))

#change data types to numeric if necessary
smart1 <- smart1 %>%
  mutate(across(where(is.character), as.numeric))

```

# Logistic Regression Model

```{r}
#| echo: TRUE
#| warnings: FALSE

model <- glm(EVENT ~ AGE + SEX + BMI + SYSTH + HDL + DIABETES + HISTCAR2 +
               HOMOC + log(CREAT) + STENOSIS + IMT + SMOKING + alcohol + albumin,
             data = smart1, family = binomial)

```

# Question A

The performance of the logistic regression model is evaluated using 5-folds cross validation (n = 3873). At the default probability threshold of 0.5, the mean sensitivity was 0.05 (SD = 0.02) and the mean specificity was 0.99 (SD = 0.05).

The mean area under the ROC curve across the folds was 0.73 (SD = 0.03), and the AUC of the ROC based on the pooled cross-validated predictions was 0.73, see Figure. This means that a randomly chosen patient with an event receives a higher predicted probability than a randomly chosen patient without an event in about 73% of cases. This is better than the random classifier (AUC = 0.5), but far from perfect discrimination (AUC = 1). Furthermore, the ROC curve shows how sensitivity and specificity trade off as the threshold varies. For example, for a sensitivity of roughly 0.8, a specificity of roughly 0.5 should be accepted.

```{r}
#| echo: TRUE
#| warnings: FALSE

# same proportion in every fold, check if that is correct
# way of validating the final model of part 1 with no external data available

library(caret)
library(pROC)

set.seed(1234)

folds <- createFolds(smart1$EVENT, k = 5, returnTrain = TRUE) # ensures by default that EVENT proportions are equal 
print(str(folds))

train_control <- trainControl(
  method = "cv",
  number = 5,
  index = folds,
  classProbs = TRUE,                  # needed to get predicted probabilities, not just class labels
  summaryFunction = twoClassSummary,  # computes ROC/AUC, sensitivity, specificity
  savePredictions = "final"           # keeps held-out predictions for plotting the ROC curve
)

model_glm <- train(
  EVENT ~ AGE + SEX + BMI + SYSTH + HDL + DIABETES + HISTCAR2 +
    HOMOC + log(CREAT) + STENOSIS + IMT + SMOKING + alcohol + albumin,
  data = smart1,
  method = "glm",
  family = "binomial",
  trControl = train_control,
  metric = "ROC"                      # tells caret to select/report based on AUC
)

# View CV performance
print(model_glm)
model_glm$results     # mean ROC (AUC), Sens, Spec across the 5 fold (sensitivity and specificity are flipped due to sequence in labeling)

# Extract the saved CV predictions
cv_preds <- model_glm$pred

# Confusion matrix at 0.5, using the labels caret stored
confusionMatrix(cv_preds$pred, cv_preds$obs, positive = "yes")  # sensitivity for events

# Build ROC object (probability of "yes" vs actual class)
roc_obj <- roc(response = cv_preds$obs, 
               predictor = cv_preds$yes,
               levels = c("no", "yes"))

# Plot
plot(roc_obj, 
     main = "ROC Curve (5-fold CV)", 
     col = "blue", 
     lwd = 2,
     xlim = c(1, 0),
     xaxs = "i",          
     yaxs = "i",
     xlab = "Specificity",
     ylab = "Sensitivity")

auc(roc_obj)
```

# Question B

```{r}
#| echo: TRUE
```

# Question C

Below, a comparison between the predictive performance of the logistic regression model and the Linear discriminant analysis model is made. For both analysis, the same predictors are used.

Below an Linear discriminant analysis is executed.

```{r}
#| echo: TRUE
warnings = FALSE
message = FALSE
library(caret)
library(MASS)
library(pROC)


set.seed(1234)

model_lda <- train(
  EVENT ~ AGE + SEX + BMI + SYSTH + HDL + DIABETES + HISTCAR2 +
    HOMOC + log(CREAT) + STENOSIS + IMT + SMOKING + alcohol + albumin,
  data = smart1,
  method = "lda",
  trControl = train_control,
  metric = "ROC"
)

print(model_lda)

```

Both models are validated using 5-fold-cross-validation. For each fold the area under the ROC (AUC) is determined. The mean AUC of these five values is used as estimate of the performance of the prediction of the model.

```{r}
warnings = FALSE
message = FALSE
# Mean performance across the 5 folds
model_lda$results

# Performance separately for each fold
model_lda$resample

cat("Linear discriminant analysis model: \n")
mean_lda_auc <- model_lda$results$ROC
mean_lda_auc

cat("Logistic regression model: \n")
mean_glm_auc <- model_glm$results$ROC
mean_glm_auc
```

the mean AUC of the performed cross validation of the LDA model is 0.733 and 0.732 for the logistic regression model. So, the difference in AUC between the LDA and the logistic regression model is very small. However, this does not take uncertainty into account. Below we will find the 95% confidence interval of the AUC for both models.

```{r}
warnings = FALSE
message = FALSE
glm_preds_compare <- model_glm$pred %>%
  dplyr::select(
    rowIndex,
    Resample,
    obs,
    glm_prob = yes
  )

lda_preds_compare <- model_lda$pred %>%
  dplyr::select(
    rowIndex,
    Resample,
    obs,
    lda_prob = yes
  )

compare_preds <- inner_join(
  glm_preds_compare,
  lda_preds_compare,
  by = c("rowIndex", "Resample", "obs")
)

```

```{r}

roc_glm_compare <- roc(
  response = compare_preds$obs,
  predictor = compare_preds$glm_prob,
  levels = c("no", "yes")
)

roc_lda_compare <- roc(
  response = compare_preds$obs,
  predictor = compare_preds$lda_prob,
  levels = c("no", "yes")
)
cat("Logistic regression model\n")
cat("AUC:", round(as.numeric(auc(roc_glm_compare)), 4), "\n")
cat(
  "95% CI:",
  round(ci.auc(roc_glm_compare)[1], 4), "-",
  round(ci.auc(roc_glm_compare)[3], 4), "\n\n"
)

cat("Linear discriminant analysis model\n")
cat("AUC:", round(as.numeric(auc(roc_lda_compare)), 4), "\n")
cat(
  "95% CI:",
  round(ci.auc(roc_lda_compare)[1], 4), "-",
  round(ci.auc(roc_lda_compare)[3], 4), "\n"
)
```

The 95% confidence interval of the AUC of the pooled out-of-fold validation is calculated for both models. In this method the predictions for all 5 folds were combined and one AUC is calculated. Instead of taking the average of the 5 AUCs. Which causes a small variation between this 2 methods.

Using this out of fold method. For the logistic regression we have a estimated area under the curve of 0.7309, with a 95% confidence interval of (0.7058 ; 0.7561). For the LDA method we estimate the area under the curve 0.7304, with a 95% confidence interval of (0.7051;0.7557). So, the pooled AUCs of both models are almost identical with overlapping 95% confidence intervals. So with a 95% confidence interval there is no clear evidence of a difference between the 2 models.
